Application of Machine Learning to Increase the Efficiency of the Global Search Algorithm for Solving Multicriterial Problems\thanks{This research was supported by the Russian Science Foundation, project No. 21-11-00204.}

### 2023-10-16 2023/NUMTA/Full paper/NUMTA_2023_MCO.tex
fix NUMTA paper

As the decision-making model we consider the multicriterial multidimensional optimization problems with multiextremal ``black-box'' criteria herein after called as MMCO problems. Combination of multicriteriality and multiextremality generates the dramatic complexity of these problems, because, first, criteria are contradictory, as a rule, and this circumstance leads to a complicated concept of solution (Pareto set) in such problems and, second, in the multiextremal case the growth of computational efforts is exponential when increasing the dimensionality (number of independent variables).

@@

In the class of deterministic methods for multiextremal optimization with guaranteed convergence to global optimum \cite{Gablonsky2001,Sergeyev2010,Sergeyev2006,Evtushenko2014,Strongin2003,Sergeyev2013}, the algorithms developed in the framework of the information-statistical approach (ISA) \cite{Gergel2018,Barkalov2018} are among the most effective. The present paper describes a new algorithm belonging to the ISA class and aimed at solving MMCO problems that accelerates the Pareto set building by the application of machine learning ideas.

@@

Unfortunately, finding a decision   that provides the minimal values for all partial criteria simultaneously is possible only in exceptional cases because, as a rule, partial criteria may be contradictory, i.e. when decreasing one criterion leads to increasing the other.

@@

In general case, the functions satisfying the Lipschitz condition are multiextremal, therefore, the convolution (\ref{eq:05}) will be multiextremal and Lipschitzian as well. It means that for solving the problem (\ref{eq:08}) it is required to apply methods of nonlinear programming which provide finding the global optimum. There are many global optimization algorithms designed under different assumptions about properties of objective function; relevant references can be found, for instance, in monographs \cite{Pardalos2017,Marler2009,Sergeyev2013}. Some of them are theoretically substantiated and guarantee the convergence to global solution; the others are empirical, for example, nature-inspired methods that do not ensure finding the global optimum.

@@

The following computational scheme is proposed for solving the problems of the family (\ref{eq:15}). In this scheme, all the problems (\ref{eq:15}) are solved sequentially and during minimization of current function $f^i(x), i > 1$, the results of preceding problems $f^j(x) \to \min, 1 \leq j \leq i - 1,$ completed already are taken into account.

@@

Three series of experiments were executed. The first one dealt with a simple problem and was aimed at demonstration of positive influence of criteria space analysis by means of separating planes and at comparing with evolutionary techniques.
This problem was bi-criteria one \cite{CHIANDUSSI2012912}:
\begin{equation}
    \label{eq:28}
    \begin{matrix}
f_1 (y)=y_1, \\
f_2 (y)=(1+10y_2 )[1- \left(\frac{y_1}{1+10y_2} \right)^2- \frac{y_1}{1+10y_2} \sin(8 \pi y_1 ) ], \\
0 \leq y_1,y_2 \leq 1.
    \end{matrix}
\end{equation}

>>>>

As the decision-making model we study the multicriterial multidimensional optimization problems with multiextremal ``black-box'' criteria hereinafter called as MMCO problems. The property of multiextremality means that any criterion can have several local minima. Combination of multicriteriality and multiextremality generates the dramatic complexity of these problems, because, first, criteria are contradictory, as a rule, and this circumstance leads to a complicated concept of solution (Pareto set) in such problems and, second, in the multiextremal case the growth of computational efforts is exponential when increasing the dimensionality (number of independent variables).

@@

In the class of deterministic methods for multiextremal optimization with guaranteed convergence to global optimum \cite{Gablonsky2001,Sergeyev2010,Sergeyev2006,Evtushenko2014,Strongin2003,Sergeyev2013}, the algorithms developed in the framework of the information-statistical approach (ISA) \cite{Gergel2018,Barkalov2018} are among the most effective. In the ISA framework, the optimized function is considered as a sample of a random function with predefined stochastic properties (information), and the optimization method is designed as a statistical procedure that optimizes the probability of global minimizer finding. The present paper describes a new algorithm belonging to the ISA class and aimed at solving MMCO problems that accelerates the Pareto set building by the application of machine learning ideas.

@@

Unfortunately, finding a decision that provides the global minimal values for all partial criteria simultaneously is possible only in exceptional cases because, as a rule, partial criteria may be contradictory, i.e. when decreasing one criterion leads to increasing the other.

@@

In should be noted, that under the conditions (\ref{eq:09}), the convolution (\ref{eq:05}) is Lipschitzian as well. Moreover, this convolution is multiextremal in the case of multiextremality of partial criteria $\varphi_s(y)$, $1 \leq s \leq k$. It means that for solving the problem (\ref{eq:08}) it is required to apply methods of nonlinear programming which provide finding the global optimum. There are many global optimization algorithms designed under different assumptions about properties of objective function; relevant references can be found, for instance, in monographs \cite{Pardalos2017,Marler2009,Sergeyev2013}. Some of them are theoretically substantiated and guarantee the convergence to global solution; the others are empirical, for example, nature-inspired methods that do not ensure finding the global optimum.

@@

The following computational scheme is proposed for solving the problems of the family (\ref{eq:15}). In this scheme, all the problems (\ref{eq:15}) are solved sequentially and during minimization of current function $f^i(x), i > 1$, the results of preceding problems $f^j(x) \to \min, 1 \leq j \leq i - 1$, completed already are taken into account. The main idea is to calculate values of the function $f_i(x)$ by utilizing the values of partial criteria computed earlier and to incorporate them to the scheme of the global search algorithm as results of initial iterations.

@@

Three series of experiments were executed. The first one dealt with a simple problem and was aimed at efficiency demonstration of the proposed technique that combines the ISA method with separating planes (AGS-SP) compared to its prototype AGS-R and evolutionary algorithms.
This problem was bi-criteria one \cite{CHIANDUSSI2012912}:
\begin{equation}
    \label{eq:28}
    \begin{matrix}
f_1 (y)=y_1, \\
f_2 (y)=(1+10y_2 )[1- \left(\frac{y_1}{1+10y_2} \right)^2- \frac{y_1}{1+10y_2} \sin(8 \pi y_1 ) ], \\
0 \leq y_1,y_2 \leq 1.
    \end{matrix}
\end{equation}
